\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-06-27 eg6-letters/NIM/EG6_RTPC_NIM.tex
Add Stepan's comments

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
Laboratory to track low-energy nuclear recoils for the purpose of measuring
exclusive nuclear channels, such as coherent Deeply Virtual Compton Scattering
and coherent meson production on $^4$He. In such processes, the $^4$He nucleus
remains intact in the final state, however the CEBAF Large
Acceptance Spectrometer (CLAS) cannot track the low energy $\alpha$ particles. In
2009, we carried out measurements using the CLAS spectrometer supplemented by
the RTPC positioned directly around a gaseous $^4$He target, allowing a detection
threshold as low as 300~MeV/c for $^4$He. This article discuses the design,
work principle, calibration methods and performances of this RTPC.
\end{abstract}

@@

Until recently, Thomas Jefferson National Accelerator Facility (TJNAF), in 
Newport News, Virginia, USA, has provided high power electron beams of energy 
up to 6 GeV and 100$\%$ duty cycle to three experimental Halls (A, B, C) 
simultaneously. The CLAS spectrometer~\cite{CLASref}, located
in Hall-B, is composed of several sub-detectors and two magnets. 
Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
    particles~\cite{DCref} (in blue on the figure).
 \item Superconducting toroidal magnet (in yellow) to bend the trajectories 
    of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Cerenkov Counters (CC) to distinguish electrons from other negative 
    particles, such as pions, at lower energies~\cite{CCref} (in pink).
 \item Scintillation Counters (SC) to identify hadrons by measuring their 
    time of flight (TOF)~\cite{TOFref} (in red).
 \item Electromagnetic Calorimeters (EC) to distinguish electrons from other negative 
    particles at higher energies and to detect neutral particles (neutrons and 
    photons in particular)~\cite{ECref} (in green).
\end{itemize}

@@

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a nucleon or nuclei) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005~\cite{FX}.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long, corresponding
to 17 radiation lengths. The achieved energy resolution is around 3 to 4$\%$ for photon
energies between 2 GeV and 5 GeV, with an angular resolution between 3 and 5 mrad 
\cite{Hyon-suk}. To protect this detector and CLAS from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the IC. 
In addition, the recoil nuclei from the coherent DVCS on helium was detected 
using a RTPC developed to track low energy nuclear fragments. The solenoid field
serving as the analyzing magnet for particle tracking in the RTPC. This 
setup was used during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 250 MeV/c. Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fitting within the 230~mm diameter 
of the 5 Tesla solenoid magnet surrounding the target.

@@

The new CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electric field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of curved gas electron multipliers (GEM) and 
detected by the readout system 
on the external shell of the detector as illustrated in the 
Fig.~\ref{fig:RTPC_1_4}. The RTPC is segmented into two halves with 
independent GEM amplification systems\footnote{The foils were produced by 
Tech-Etch, Inc.} that cover about 80\% of the azimuthal 
angle.

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item First, the 6 atm helium-4 target extends along the beam line axis; it 
      is 284~mm long with a 3~mm radius wall made of a 27-$\mu$m-thick Kapton.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      that is electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away 
      from the beam axis. The average electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower Voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

The GEM technology has been chosen for the flexibility of the GEM foils,
which can be easily used to produce a curved amplification surface. Also, 
GEMs are known to have relatively low sparks rate~\cite{GEM_ref}, which 
is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 50~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in Fig.~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

@@

One of the important challenge in developing the radial TPC was how to obtain a 
good support structure for the GEM foils to allow a tractable 
installation of the GEMs. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each end of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations. The rigidity of the GEM foils was enough
for the structure to be self-supporting and only the backward part of the 
cylinder was fixed to the main mechanical structure. This design only left a 
light fiberglass ring in the forward region, reducing to a minimum secondary 
interactions.

@@

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{MAGBOLTZ} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, each iteration increasing the number of events 
properly reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this long 
calibration process.

@@

We determine the drift speed using tracks reconstructed in the RTPC. In 
Fig.~\ref{fig:RTPC_signals}, a typical $^{4}$He track is represented (in green). After 
it causes ionization in the drift region, the released electrons (in black) 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, but cylindrical symmetry insures they always travel the same 
distance. By identifying the maximum time measured ($T_{Max}$), we can infer the drift 
speed of the electrons in the RTPC.\\

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate the drift paths, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate directly the Helium nucleus one. 
By comparing this calculated kinematic to the measurement in the RTPC, we fine tuned
the drift paths independently of our knowledge of the exact chamber's conditions.

The drift paths are adjusted using a set of identified elastic events from our 
lower beam energy run. Based on the kinematic of electrons in these events, 
we generate the helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then 
we compare the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber.

Because of the magnetic field, the drift paths are not linear in the RTPC. So 
to perform the extraction, we make a first approximation with a linear 
dependence between the radius of emission and the time of detection, and then 
refine our result. As it happens, the curvature is minimal and this process 
converges already on the second iteration.

@@

The gain of each pad is defined as the ratio between the actual 
deposited energy and the registered ADC value. We tested two different 
methods to extract such gains, with varying degrees of success.

The first method is based on the measured average $\small{\frac{dE}{dX}}$
and the expected value based on the Bethe-Bloch formula. The total energy
deposition for each $^4$He track is calculated as the sum of all ADC hits
attributed to the track. The path length of that energy deposition is
calculated as the distance along the reconstructed trajectory between the first
and last hits on the track and corrected to account for regions
of bad channels along the track. The measured average energy loss is then the
ratio of deposited energy and path length. The Bethe-Bloch formula is
evaluated at the track's reconstructed momentum, and the ratio with the measured
$\small{\frac{dE}{dX}}$ is taken as a gain scaling factor. This average gain is
attributed equally to each readout channel contributing to the track, and each
channel's gain is averaged over many tracks. This gain calibration method is
inherently iterative, and, while it does result in measured $\small{\frac{dE}{dX}}$
much closer to the expected Bethe-Bloch curve than prior to calibration, the
results shown in Fig.~\ref{fig:dedx_p_exp_1st} are not very satisfactory.

\begin{figure*}[tb]
   \centering
   \includegraphics[scale=0.73]{fig_2017/f_dedx_p_exp_1st.pdf}
   \caption{$\small{\frac{dE}{dX}}$ vs. $p/q$ distribution for the left half of 
      the RTPC (left) and for the right half (right). Here, 
      $\small{\frac{dE}{dX}}$ is calculated using the gains of the first 
      method.  The lines are theoretical expectations from Bethe-Bloch formula 
   for $^4$He (red), $^3$He (black), $^3$H (blue), $^2$H (pink) and protons (grey).}
\label{fig:dedx_p_exp_1st}
\end{figure*}

The second method is to compare the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. In this step, the gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. 
Results of this second method are shown in Fig.~\ref{fig:dedx_p_exp_2nd},
from these we concluded that the gains of the 
second method match best the theoretical lines.

@@

\begin{figure*}[tb]
\centering
\includegraphics[scale=0.73]{fig_2017/f_dedx_p_exp_2nd.pdf}
\caption{$\small{\frac{dE}{dX}}$ vs. $p/q$ distribution for the left half of 
   the RTPC (left) and for the right half (right). Here, 
   $\small{\frac{dE}{dX}}$ is calculated using the gains of the second method.  
   The lines are theoretical expectations from Bethe-Bloch formula for $^4$He 
   (red), $^3$He (black), $^3$H (blue), $^2$H (pink) and protons (grey).}
\label{fig:dedx_p_exp_2nd}
\end{figure*}

>>>>

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
Laboratory to track low-energy nuclear recoils for the purpose of measuring
exclusive nuclear reactions, such as coherent Deeply Virtual Compton Scattering
and coherent meson production on $^4$He. In such processes, the $^4$He nucleus
remains intact in the final state, however the CEBAF Large
Acceptance Spectrometer (CLAS) cannot track the low energy $\alpha$ particles. In
2009, we carried out measurements using the CLAS spectrometer supplemented by
the RTPC positioned directly around a gaseous $^4$He target, allowing a detection
threshold as low as 12~MeV for $^4$He. This article discuses the design,
work principle, calibration methods and the performance of this RTPC.
\end{abstract}

@@

Until recently, Thomas Jefferson National Accelerator Facility, in 
Newport News, Virginia, USA, has provided high power electron beams of 
up to 6 GeV energy and 100$\%$ duty cycle to three experimental Halls (A, B, C) 
simultaneously. The CLAS spectrometer~\cite{CLASref}, located
in Hall-B, is based on a superconducting toroidal magnet and composed of 
several sub-detectors. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
       particles~\cite{DCref} (in blue on the figure).
 \item Superconducting toroidal magnet (in yellow) to bend the trajectories 
       of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Threshold Cherenkov Counters (CC) for electron identification at momenta 
       $<2.7$ GeV~\cite{CCref} (in pink).
 \item Scintillation Counters (SC) to identify charged hadrons by measuring their 
       time of flight (TOF)~\cite{TOFref} (in red).
 \item Electromagnetic Calorimeters (EC) for identification of electrons, 
       photons and neutrons~\cite{ECref} (in green).
\end{itemize}

@@

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a nucleon or nuclei) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005~\cite{FX}.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long, corresponding
to 17 radiation lengths. The achieved energy resolution is around 3 to 4$\%$ for photon
energies between 2 GeV and 5 GeV, with an angular resolution between 3 and 5 mrad 
\cite{Hyon-suk}. To protect the CLAS detector and the IC from the large flux 
of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on helium, a new 
radial time projection chamber was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in RTPC. This 
setup was used during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 300 MeV/c (12 MeV kinetic energy). Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fit inside the 230 mm diameter bore of the 
5 Tesla solenoid magnet.

@@

The new CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electric field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of cylindrical gas electron multipliers (GEM) and 
detected by the readout system 
on the external shell of the detector as illustrated in the 
Fig.~\ref{fig:RTPC_1_4}. The RTPC is segmented into two halves with 
independent GEM amplification systems\footnote{The foils were produced by 
Tech-Etch, Inc.} that cover about 80\% of the azimuthal 
angle.

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item The 6 atm helium-4 target extends along the beam line axis; it 
      is a 284~mm long, 6~mm in diameter straw with a 27-$\mu$m Kapton wall.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      that is electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away 
      from the beam axis. The average electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower Voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

The GEM technology has been chosen for the flexibility of the GEM foils,
which can be easily used to produce a curved amplification surface. Also, 
GEMs are known to have relatively low spark rate~\cite{GEM_ref}, which 
is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 300~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in Fig.~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

@@

One of the important challenge in developing the radial TPC was how to obtain a 
good support structure for the GEM foils to allow a tractable 
installation of the GEMs. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each end of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations. The rigidity of the GEM foils was enough
for the structure to be self-supporting and only the upstream end of the cylinder 
was fixed to the main mechanical structure. This design only left a 
light fiberglass ring in the downstream end, reducing to a minimum secondary 
interactions.

@@

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{MAGBOLTZ} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, each iteration increasing the number of events 
properly reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

After the ionization, the released electrons 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, but cylindrical symmetry insures that they always travel the same 
distance. We illustrate this in Fig.~\ref{fig:RTPC_signals}, where we depict a 
typical $^{4}$He track. By measuring the maximum time ($T_{Max}$), we can infer the drift 
speed of the electrons in the RTPC.\\

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate them, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate directly the kinematic of the Helium nucleus. 
By comparing the alpha particle's calculated momentum and angle of recoil to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the exact chamber's conditions.

Based on the kinematic of electrons in the calibration data, 
we generate the helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then 
we compare the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber. 
Because of the magnetic field, the drift paths are not linear in the RTPC. So 
to perform the extraction, we make a first approximation with a linear 
dependence between the radius of emission and the time of detection, and then 
refine our result. As it happens, the curvature is minimal and this process 
converges already on the second iteration.

@@

The gain of each pad is defined as the ratio between the actual 
deposited energy and the registered ADC value. 
To do so we compare the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}.

@@

\begin{figure*}[tb]
\centering
\includegraphics[scale=0.73]{fig_2017/f_dedx_p_exp_2nd.pdf}
\caption{$\small{\frac{dE}{dX}}$ vs. $p/q$ distribution for the left half of 
   the RTPC (left) and for the right half (right) after gain calibration.  
   The lines are theoretical expectations from Bethe-Bloch formula for $^4$He 
   (red), $^3$He (black), $^3$H (blue), $^2$H (pink) and protons (grey).}
\label{fig:dedx_p_exp_2nd}
\end{figure*}
